Rust for Scientific Computing

### 2025-09-24 paper/main.tex
more notes, more formatting, more lazy commit messages

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[hidelinks]{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{layouts}
\usepackage{microtype}      % microtypography
\usepackage{mathrsfs}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{graphicx}
\usepackage{doi}
\usepackage{listings}
\usepackage{tikz}
\usepackage[backend=biber,style=ieee]{biblatex}

@@

@@

\twocolumn
\pagebreak

\printinunitsof{in}\prntlen{\columnwidth}

% keywords can be removed
\keywords{Rust \and HPC \and Scientific Computing \and Programming}

@@

Programming languages sit at the heart of all things computation.
They allow users to express higher-level instructions in a human-readable fashion such that this
code can be translated into machine-level instructions by either compiling or interpreting it.
Various classes of programming languages have emerged that serve distinct purposes.
Fundamentally, they aim to descibe problems at various levels of depth which allows humans to
interpret and reason about the executed instructions and therefore express their desired actions
more clearly.

Importance of programming in science
\begin{enumerate}
    \item New fields of study (computational physics, chemistry, biology, sociology, etc.)
    \item Cite some groundbreaking examples which were possible due to programming in
        science (maybe weather)
    \item
\end{enumerate}

Many languages have been devised to target specific realms which have emerged over the past decades.
They span across a range of abstraction levels, ranging from low-level languages to high-level ones.
The low-level operations level mostly deals with instructions on a level close to the actual
hardware such as memory allocation, pointer arithmetic or by using assembly instructions directly.
Meanwhile, high-level operations usually combine many lower-level ones to perform a single task.
An example for a high-level operation would be the addition of two matrices which might allocate the
required memory, fill it according to the given values and then call the appropriate math function
to pairwise add their entries.
Languages which can operate effectively on a lower level often bear the most promise for maximum
performance which is crucial in any scientific application.
These languages are mostly compiled using an optimizing compiler which enables many of their
performance achievements.

This mini-review concerns itself with the Rust programming language and aims to provide an overview
of the field of scientific computing with the Rust ecosystem.
Other systems programming languages  such as C, C++ or Fortran have dominated the field over the
past decades.
Most notably, C++ has been influential in establishing a model of "zero cost abstractions" where a
phrasing of the problem from a higher point of view does not involve any additional execution cost
but is handled by the compiler.
These and many other principles have allowed for a plethora of

Judging C/C++
\begin{enumerate}
    \item Previous Decades: C/C++ in science
    \item Limitations of design choices of language become evident
    \item fragmented ecosystem; good for competition; bad for developer experience
    \item modern features missing: package manager with package registries, build system
    \item technical details; mutable by default, templates vs generics, backwards
        compatiblity becomes a
        problem ("there is a much simpler langauge within C++ that tries to get out"),
        macros, no clear
        ownership rules on a language level, dangling pointers
    \item still many good principles (such as RAII) and design principles/concepts (cite
        gang of four)
\end{enumerate}

@@

Show that many companies are starting to adopt Rust
\begin{enumerate}
    \item Amazon (AWS)\\ AWS developed Firecracker, a microVM monitoring service powering
        AWS Lambda and AWS
        Fargate. Firecracker is built primarily in Rust to offer fast cold boot times and
        secure workload
        isolation.
    \item Rust-supported runtimes like AWS Lambda also allow developers to write
        serverless functions using
        Rust.
    \item Google\\ Fuchsia, Google's modern operating system, uses Rust extensively for
        system components
        such as drivers and services to enhance security and reliability.
    \item Google (Android)\\ Android integration: Rust has been adopted within the
        Android Open Source
        Project (AOSP) to improve memory safety in critical system components. DEV
        Community Litslink
    \item Meta (Facebook)\\ Mononoke, Facebook’s large-scale source control backend for
        Mercurial, was
        written in Rust to improve performance and reliability in its monorepo.
    \item Meta (Facebook)\\ Facebook employs Rust in developing CLI tools, distributed
        systems, and cron
        jobs. Engineers write millions of lines of Rust internally.
    \item Apple\\ Apple’s Cloud Traffic team and backend infrastructure have been
        shifting from C to Rust,
        with new functionality being primarily built in Rust. This change was driven by performance,
        security, and developer efficiency goals. Reddit
    \item Microsoft\\ Although not part of FAANG, it’s noteworthy—Microsoft is rewriting
        parts of the Windows
        kernel and system components in Rust (e.g., DWriteCore, GDI Regions),
        prioritizing memory safety.
    \item Azure services-like Azure IoT Edge—also incorporate Rust.
    \item Linux Kernel\\ Only other supported language except from C
\end{enumerate}

\begin{figure*}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-popular-languages.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/pypi-org-used-languages.pdf}
    \caption{
        Usage of programming languages according to (A) stackoverflow.com and (B) pypi.org.
    }
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-loved-language.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-desired-language.pdf}
    \caption{
        Popularity of the Rust programming language.
    }
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/crates-io-release-numbers.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/crates-io-deps-downloads-scatter.pdf}
    \caption{
        Number of releases of all included crates.
        Measure of overall growth of the field.
        The second figure can be misleading since this will for example not capture the setup where
        a crate is the dependency of a pypi.org package which is downloaded very often.
        The colors and size of the markers are also scaled logarithmically with the number of total
        downloads.
    }
\end{figure*}

@@

\begin{table*}
    \centering
    \begin{tabular}{l r r r r r}
        \toprule
        Crate Name &Last Update &Version &$N_d$ &$D_\text{week}$ &$D_\text{total}$\\
        \midrule
        Lamellar         &2025-03-31 &0.7.1      &42   &20       &     16.9k\\
        argmin           &2024-02-27 &0.10.0     &18   &12003    &   1227.7k\\
        arrow            &2025-08-25 &56.1.0     &23   &7532     &  27425.0k\\
        bevy             &2025-05-30 &0.16.1     &35   &10175    &   2847.7k\\
        bless            &2025-07-20 &0.1.1      &13   &80       &      0.6k\\
        burn             &2025-07-18 &0.18.0     &12   &1793     &    319.7k\\
        c3dio            &2024-04-08 &0.8.0      &2    &24       &     11.8k\\
        cargo-upgrades   &2025-08-25 &2.2.4      &10   &553      &     64.7k\\
        cellular\_raza   &2025-08-05 &0.3.0      &10   &157      &     27.4k\\
        cfsem            &2025-08-14 &2.5.0      &8    &106      &      4.9k\\
        city2ba          &2020-02-28 &1.1.0      &20   &19       &      3.0k\\
        cmake            &2025-02-10 &0.1.54     &1    &899615   &  91211.9k\\
        cudarc           &2025-08-07 &0.17.2     &5    &1184     &    647.0k\\
        cxxbridge        &2019-12-28 &0.0.0      &0    &14       &      1.3k\\
        deimos           &2025-06-23 &0.10.0     &13   &40       &      7.1k\\
        diffsol          &2025-07-16 &0.6.6      &16   &131      &     33.9k\\
        dioxus           &2025-08-11 &0.6.3      &28   &         &    637.5k\\
        egobox           &2025-08-22 &0.32.0     &18   &250      &     35.6k\\
        enzyme           &2021-12-26 &0.4.0      &5    &17       &      5.4k\\
        extendr-api      &2025-07-12 &0.8.1      &11   &968      &    178.4k\\
        factrs           &2025-01-13 &0.2.0      &20   &24       &      1.8k\\
        faer             &2025-04-29 &0.22.6     &46   &1322     &    528.4k\\
        faucet           &2025-05-31 &0.2.0      &7    &33       &      3.0k\\
        fem\_2D          &2023-04-20 &0.2.2      &6    &20       &      5.4k\\
        gcv\_spline      &2024-01-28 &0.2.0      &1    &16       &      2.5k\\
        \bottomrule
    \end{tabular}
    \caption{TODO}
\end{table*}

@@

\onecolumn
\printbibliography

>>>>

\usepackage{acronym}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[hidelinks]{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{layouts}
\usepackage{microtype}      % microtypography
\usepackage{makecell}
\usepackage{mathrsfs}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{pagecolor}
\usepackage{graphicx}
\usepackage{doi}
\usepackage{listings}
\usepackage{tikz}
\usepackage{tabularx}
\usepackage[backend=biber,style=ieee]{biblatex}
\usepackage[table]{xcolor}

@@

\newacro{raii}[RAII]{Resource Acquisition is Initialization}
\newacro{ode}[ODE]{Ordinary Differential Equation}

@@

% \twocolumn
\pagebreak

% keywords can be removed
\keywords{Rust \and Scientific Computing \and High-Performance Computing \and Programming}

@@

Programming languages sit at the heart of all things computation.
They allow users to express higher-level instructions in a human-readable fashion such that this
code can be translated into machine-level instructions by either compiling or interpreting it.
Various classes of programming languages have emerged that serve distinct purposes.
Fundamentally, they aim to descibe problems from different perspectives which allows humans to
interpret and reason about the executed instructions and therefore express their desired actions
more clearly.

Languages such as C, C++ or Fortran have dominated the field of compiled languages over the past
decades.
These languages are able to solve many different problems in a general-purpose fashion.
Early achievements of the first commercially available programming language Fortran are still used
today in the Modular Ocean Model~\cite{Adcroft2019,Griffies2012}.
The development of the C programming language in 1972 enabled milestones such as the Unix operating
system~ \cite{Ritchie1978} which layed the foundation for many major scientific achievements.
With the advent of C++ in 1985 further contributions were made such as the root project which
is an "Object Oriented Data-Analysis Framework"~\cite{Brun1997} and mainly used by
CERN~\cite{https://doi.org/10.7483/opendata.atlas.l806.5cku}.
They also build the backbone of all deep learning tools such as
pytorch~\cite{Ansel_PyTorch_2_Faster_2024} and tensorflow~\cite{tensorflow2015-whitepaper} which
have seen widespread use and are written on a higher abstraction level using Python.
Tensorflow builds the foundation of AlphaFold~\cite{Jumper2021} which provides AI-based protein
structure prediction and Pytorch enabled the development of new cell segmentation
tools~\cite{Cutler2022}.

% Importance of programming in science
% \begin{enumerate}
%     \item New fields of study (computational physics, chemistry, biology, sociology, etc.)
%     \item Cite some groundbreaking examples which were possible due to programming in
%         science (maybe weather)
%     \item
% \end{enumerate}

A single programming language can span across multiple abstraction levels, ranging from low-levels
of operation to high-level ones.
The former mostly deal with instructions close to the actual hardware such as memory allocation,
pointer arithmetic or by using assembly instructions directly.
Meanwhile, high-level operations usually combine many lower-level ones to perform a single task.
An example for a high-level operation would be the addition of two matrices which might allocate the
required memory, fill it according to the given values and then call the appropriate math function
to pairwise add their entries.
Languages which can operate effectively on a lower level often bear the most promise for maximum
performance which is crucial for most scientific applications.
% Of these languages almost all are compiled using an optimizing compiler which enables many of their
% performance achievements.

This review concerns itself with the Rust programming language and aims to provide an overview
of the field of scientific computing within the Rust ecosystem.
Rust supports multiple programming paradigms and offers built-in memory safety by employing an
ownership model that enforces the validity of all references.
As such it is able to cover a wide range of abstraction levels, comparable with C++ and offers
comparable performance \textbf{CITATION}.
It takes hard-learned lessons and offers solutions for many existing problems that other languages
face.
It has been the "most loved language" for the past 7 years according to data from the developer
survey of \url{stackoverflow.com}.
This data is shown in figure~\ref{fig:stackoverflow-popular-languages} where we compare it to other
popular programming languages is compared against other languages.
It also ranks high among the "most desired languages" which is displayed in
figure~\ref{fig:stackoverflow-desired-languages}.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-loved-language.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-desired-language.pdf}
    \caption{
        Popularity of the Rust programming language.
    }
    \label{fig:stackoverflow-desired-languages}
\end{figure}

\subsection{Comparison of Rust with C and C++}
Within the realm of scientific computing, C++ and C can be seen as the major competitors of Rust.
As previously discussed, they have been instrumental in the achievement of many scientific
milestones.
They are widely applied and developers are generally readily available to tackle new projects.
Thus, their importance to the foundational advances across all areas of computation should not be
understated.
However, as with any maturing technology, many issues have been identified which are intrinsic to
the design of these languages.
The C programming language manually manages the memory of the program which provides flexibility but
leaves the task of correctly freeing resources to the user which can quickly become difficult to
track.
This ambiguity has led to many bugs and vulnerabilities.
The chromium project has published data that indicates $\approx 70\%$ of all high-severity bugs to
be related to memory unsafety~\cite{ChromiumMemorySafety}.
These issues are important for all practical applications, especially at larger scales with
multiple workstations, clusters or supercomputers where program crashes can lead to lost data and
waisted resources.

Introduce \ac{raii}

\begin{table}
    \centering
    \begin{tabularx}{\textwidth}{X X X X}
        % \toprule
        Category & C & C++ & Rust\\
        \midrule
        Memory Safety & Dangling pointers, buffer overflows, manual memory management & Includes C,
        mitigated by \ac{raii} \& smart pointers & Ownership model\\
        \midrule
        % Undefined Behavior & 
        Error Handling & Return error codes & Return error codes \& Exceptions &
        Errors are return types, \texttt{Result} and \texttt{Option} types enforce handling\\
        \midrule
        \makecell[l]{Concurrency \&\\ Parallelism} & Manual (\texttt{pthreads})
        & Manual, risk data races &
        Thread-safety through ownership model, enforced at compile-time\\
        \midrule
        Ecosystem & Extensive, fragmented & Vast, fragmented (CMake, Conan, vcpkg, ..) & Unified
        package manager, registry, build system\\
        % \bottomrule
    \end{tabularx}
\end{table}

Since C++ provides backwards compatiblity

C++ has been influential in establishing a model of "zero cost abstractions" where a phrasing of the
problem from a higher point of view does not involve any additional execution cost but is handled by
the compiler.

\begin{enumerate}
    \item Memory Safety Issues
    \item Mutable by Default
    \item Macros
    \item Ecosystem Fragmentation
    \item Concurrency and Parallelism
    \item Error Handling
\end{enumerate}

C++
\begin{enumerate}
    \item Legacy and Backward Compatibility
    \item Templates vs. Generics
    \item Complexity of the Standard
    \item Multiple Inheritance Complexity
    \item Name Lookup and Scope Complexity
    \item Undefined Behavior is Easier to Trigger
\end{enumerate}

\begin{enumerate}
    \item Previous Decades: C/C++ in science
    \item Limitations of design choices of language become evident
    \item fragmented ecosystem; good for competition; bad for developer experience
    \item modern features missing: package manager with package registries, build system
    \item technical details; mutable by default, templates vs generics, backwards
        compatiblity becomes a
        problem ("there is a much simpler langauge within C++ that tries to get out"),
        macros, no clear
        ownership rules on a language level, dangling pointers
    \item still many good principles (such as RAII) and design principles/concepts (cite
        gang of four)
\end{enumerate}

@@

\subsection{Current Advancements of Rust}
Show that many companies are starting to adopt Rust
\begin{enumerate}
    \item Amazon (AWS)\\ AWS developed Firecracker, a microVM monitoring service powering
        AWS Lambda and AWS
        Fargate. Firecracker is built primarily in Rust to offer fast cold boot times and
        secure workload
        isolation.
    \item Rust-supported runtimes like AWS Lambda also allow developers to write
        serverless functions using
        Rust.
    \item Google\\ Fuchsia, Google's modern operating system, uses Rust extensively for
        system components
        such as drivers and services to enhance security and reliability.
    \item Google (Android)\\ Android integration: Rust has been adopted within the
        Android Open Source
        Project (AOSP) to improve memory safety in critical system components. DEV
        Community Litslink
    \item Meta (Facebook)\\ Mononoke, Facebook’s large-scale source control backend for
        Mercurial, was
        written in Rust to improve performance and reliability in its monorepo.
    \item Meta (Facebook)\\ Facebook employs Rust in developing CLI tools, distributed
        systems, and cron
        jobs. Engineers write millions of lines of Rust internally.
    \item Apple\\ Apple’s Cloud Traffic team and backend infrastructure have been
        shifting from C to Rust,
        with new functionality being primarily built in Rust. This change was driven by performance,
        security, and developer efficiency goals. Reddit
    \item Microsoft\\ Although not part of FAANG, it’s noteworthy—Microsoft is rewriting
        parts of the Windows
        kernel and system components in Rust (e.g., DWriteCore, GDI Regions),
        prioritizing memory safety.
    \item Azure services-like Azure IoT Edge—also incorporate Rust.
    \item Linux Kernel\\ Only other supported language except from C
\end{enumerate}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-popular-languages.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/pypi-org-used-languages.pdf}
    \caption{
        Usage of programming languages according to (A) stackoverflow.com and (B) pypi.org.
        % For discussion: overlap between C and C++ is about 406 projects
        % Other overlaps are negligible
    }
    \label{fig:stackoverflow-popular-languages}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/crates-io-release-numbers.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/crates-io-deps-downloads-scatter.pdf}
    \caption{
        Number of releases of all included crates.
        Measure of overall growth of the field.
        The second figure can be misleading since this will for example not capture the setup where
        a crate is the dependency of a pypi.org package which is downloaded very often.
        The colors and size of the markers are also scaled logarithmically with the number of total
        downloads.
    }
\end{figure}

@@

\begin{table}
    \centering
    \rowcolors{2}{white}{lightgray!30}
    \begin{tabular}{l r r r r r}
        \toprule
        Crate Name &Last Update &Version &$N_d$ &$D_\text{week}$ &$D_\text{total}$\\
        \midrule
        Lamellar         &2025-03-31 &0.7.1      &42   &20       &     16.9k\\
        argmin           &2024-02-27 &0.10.0     &18   &12003    &   1227.7k\\
        arrow            &2025-08-25 &56.1.0     &23   &7532     &  27425.0k\\
        bevy             &2025-05-30 &0.16.1     &35   &10175    &   2847.7k\\
        bless            &2025-07-20 &0.1.1      &13   &80       &      0.6k\\
        burn             &2025-07-18 &0.18.0     &12   &1793     &    319.7k\\
        c3dio            &2024-04-08 &0.8.0      &2    &24       &     11.8k\\
        cargo-upgrades   &2025-08-25 &2.2.4      &10   &553      &     64.7k\\
        cellular\_raza   &2025-08-05 &0.3.0      &10   &157      &     27.4k\\
        cfsem            &2025-08-14 &2.5.0      &8    &106      &      4.9k\\
        city2ba          &2020-02-28 &1.1.0      &20   &19       &      3.0k\\
        cmake            &2025-02-10 &0.1.54     &1    &899615   &  91211.9k\\
        cudarc           &2025-08-07 &0.17.2     &5    &1184     &    647.0k\\
        cxxbridge        &2019-12-28 &0.0.0      &0    &14       &      1.3k\\
        deimos           &2025-06-23 &0.10.0     &13   &40       &      7.1k\\
        diffsol          &2025-07-16 &0.6.6      &16   &131      &     33.9k\\
        dioxus           &2025-08-11 &0.6.3      &28   &         &    637.5k\\
        egobox           &2025-08-22 &0.32.0     &18   &250      &     35.6k\\
        enzyme           &2021-12-26 &0.4.0      &5    &17       &      5.4k\\
        extendr-api      &2025-07-12 &0.8.1      &11   &968      &    178.4k\\
        factrs           &2025-01-13 &0.2.0      &20   &24       &      1.8k\\
        faer             &2025-04-29 &0.22.6     &46   &1322     &    528.4k\\
        faucet           &2025-05-31 &0.2.0      &7    &33       &      3.0k\\
        fem\_2D          &2023-04-20 &0.2.2      &6    &20       &      5.4k\\
        gcv\_spline      &2024-01-28 &0.2.0      &1    &16       &      2.5k\\
        \bottomrule
    \end{tabular}
    \caption{TODO}
\end{table}

@@

% \onecolumn
\printbibliography
